### 2012-11-07 paper/isca40format/AbstractProtocol.tex
will be removing evict req

\definecolor{trig-color}{gray}{.6}

@@

\begin{figure}
\begin{tabular}{|l|p{.64\columnwidth}|}
\hline
isPresent(a) & Check if cache line for address $a$ is present\\
\hline
data(a) & Cache line data for address $a$\\
\hline
state(a) & Coherence state for address $a$\\
\hline
dir(a, c) & Directory's copy of the coherence state for address $a$ in child
$c$\\
\hline
isCWait(c, a) & Check if the node waiting for a response from child $c$
for address $a$\\
\hline
cWait(c, a) & The coherence state to which the node has requested child $c$ to
go to, for address $a$\\
\hline
isPWait(a) & Check if the node waiting for a response from its parent for
address $a$\\
\hline
replacingAddr(a) & The address of the line that is going to replace the line
corresponding to address $a$\\
\hline
whoTrigger(a) & The source of the request which made the node wait for a
response. It is $None, Parent, Child \text{ or } Evict$\\
\hline
childTrigger(a) & If $whoTrigger(a)$ is $Child$, the index of the child which
is the source of the request\\
\hline
\end{tabular}
\caption{Information that the node keeps for a cache line}
\label{table:states}
\end{figure}

@@

\begin{figure}
\begin{algorithmic}
\Rule{handleCReq}{\three{ReqFromC}{c}{a}{x}}
  \If {$isPresent(a)$}
     \If {$!isPWait(a)\ \AndAnd \forall c, !isCWait(c, a)$}
        \If {$state(a) \ge x \AndAnd$ \\
              \hspace{2.5cm} $\forall c' \neq c, dir(c', a) \le \text{maxCompat}(x)$}
           \State remove$(\three{ReqFromC}{c}{a}{x})$
           \If {$dir(c,a) \neq \Inv)$}
              \State send$(\three{DataToC}{c}{a}{data(a)})$
           \EndIf
           \State $dir(c, a) \assigns x$
           \State send$(\three{RespToC}{c}{a}{x})$
        \Else
           \If {$state(a) < x$}
              \State $isPWait(a) \assigns \True$
              \State send$(\two{ReqToP}{a}{x})$
           \EndIf
           \ForAll {$c' \neq c$}
              \If {$dir(c', a) > \text{maxCompat}(x)$}
                 \State $isCWait(c', a) \assigns \True$
                 \State $cWait(c', a) \assigns \text{maxCompat}(x)$
                 \State send$(\three{ReqToC}{c'}{a}{\text{maxCompat}(x)})$
              \EndIf
           \EndFor
           \State \textcolor{trig-color}{$whoTrigger(a) \assigns Child$}
           \State \textcolor{trig-color}{$childTrigger(a) \assigns c$}
        \EndIf
     \EndIf
  \Else
     \State \textbf{let} $a' \assigns \text{getReplace}(a)$
     \If {$!isPWait(a') \AndAnd !isCWait(a')$}
        \State $replacingAddr(a') \assigns a$
        \State send$(\One{EvictReq}{a'})$
        \State \textcolor{trig-color}{$whoTrigger(a) \assigns Evict$}
        \State \textcolor{trig-color}{$childTrigger(a) \assigns c$}
     \EndIf
  \EndIf
\EndRule
\end{algorithmic}
\caption{Handle request from child}
\label{alg:handleCReq}
\end{figure}

\begin{figure}
\begin{algorithmic}
\Rule{handleEvict}{\One{EvictReq}{a}}
   \If {$!isPresent(a) \OrOr \forall c, dir(c, a) = \Inv$}
      \State remove$(\One{EvictReq}{a})$
      \If {$state(a) \neq \Inv$}
         \State send$(\two{RespToP}{a}{\Inv})$
         \If {dataXfer$(state(a), \Inv)$}
            \State send$(\two{RespToP}{a}{data(a)})$
         \EndIf
      \EndIf
      \State $isPresent(a) \assigns \False$
      \State allocLine$(replacingAddr(a))$
      \State \textcolor{trig-color}{\textbf{trigger} handleCReq($childTrigger(a)$)}
   \Else
      \If {$\forall c, !isCWait(c, a)$}
         \ForAll{$c$}
            \If {$dir(c, a) > \Inv$}
               \State $isCWait(c, a) \assigns \True$
               \State $cWait(c, a) \assigns \Inv$
               \State send$(\three{ReqToC}{c}{a}{\Inv})$
            \EndIf
         \EndFor
      \EndIf
   \EndIf
\EndRule
\end{algorithmic}
\caption{Evict a line in address $a$}
\label{alg:handleEvict}
\end{figure}

@@

\begin{figure}
\begin{algorithmic}
\Procedure{allocLine}{$a$}
  \State $isPresent(a) \assigns \True$
  \State \textcolor{trig-color}{$whoTrigger(a) \assigns None$}
  \State $state(a) \assigns \Inv$
  \ForAll {$c$}
     \State $dir(c, a) \assigns \Inv$
  \EndFor
  \State $isCWait(a) \assigns \False$
  \State $isPWait(a) \assigns \False$
\EndProcedure
\end{algorithmic}
\caption{Allocate a new cache line for address $a$}
\label{alg:allocLine}
\end{figure}

@@

Each handler will be triggered repeatedly till the corresponding message has
been removed. In order to avoid starvation, the following two conditions must be
maintained:
\begin{enumerate}
\item Once a response is available from any source, the handler for that
response must be triggered eventually.
\item After handling a response (or an evict request) the corresponding handler
must trigger the original handler that requested the received response (or made
an eviction request). The $whoTrigger(a)$ keeps track of the original handler
for each address $a$. In case of a $Child$ trigger, the actual child
corresponding to the original handler is kept track of in $childTrigger(a)$, for
address $a$, in which case $handleCReq$ is triggered for the outstanding request
from that child. In the figures, the code corresponding to this is grayed out
\end{enumerate}

>>>>

\definecolor{trig-color}{gray}{.4}

@@

\begin{figure}
\begin{tabular}{|l|p{.64\columnwidth}|}
\hline
isPresent(a) & Check if cache line for address $a$ is present\\
\hline
data(a) & Cache line data for address $a$\\
\hline
state(a) & Coherence state for address $a$\\
\hline
dir(a, c) & Directory's copy of the coherence state for address $a$ in child
$c$\\
\hline
isCWait(c, a) & Check if the node waiting for a response from child $c$
for address $a$\\
\hline
cWait(c, a) & The coherence state to which the node has requested child $c$ to
go to, for address $a$\\
\hline
isPWait(a) & Check if the node waiting for a response from its parent for
address $a$\\
\hline
whoTrigger(a) & The source of the request which made the node wait for a
response. It is $None, Parent, Child \text{ or } Evict$\\
\hline
replacingAddr(a) & If $whoTrigger(a)$ is $Evict$, the address replacing the line
of address $a$\\
\hline
childTrigger(a) & If $whoTrigger(a)$ is $Child$, the index of the child which
is the source of the request\\
\hline
\end{tabular}
\caption{Information that the node keeps for a cache line}
\label{table:states}
\end{figure}

@@

\begin{figure}
\begin{algorithmic}
\Rule{handleCReq}{\three{ReqFromC}{c}{a}{x}}
  \If {$isPresent(a)$}
     \If {$!isPWait(a) \AndAnd \forall c, !isCWait(c, a)$}
        \If {$state(a) \ge x \AndAnd$ \\
              \hspace{2.2cm} $\forall c' \neq c, dir(c', a) \le \text{maxCompat}(x)$}
           \State remove$(\three{ReqFromC}{c}{a}{x})$
           \If {$dir(c,a) \neq \Inv)$}
              \State send$(\three{DataToC}{c}{a}{data(a)})$
           \EndIf
           \State $dir(c, a) \assigns x$
           \State send$(\three{RespToC}{c}{a}{x})$
        \Else
           \If {$state(a) < x$}
              \State $isPWait(a) \assigns \True$
              \State send$(\two{ReqToP}{a}{x})$
           \EndIf
           \ForAll {$c' \neq c$}
              \If {$dir(c', a) > \text{maxCompat}(x)$}
                 \State $isCWait(c', a) \assigns \True$
                 \State $cWait(c', a) \assigns \text{maxCompat}(x)$
                 \State send$(\three{ReqToC}{c'}{a}{\text{maxCompat}(x)})$
              \EndIf
           \EndFor
           \State \textcolor{trig-color}{$whoTrigger(a) \assigns Child$}
           \State \textcolor{trig-color}{$childTrigger(a) \assigns c$}
        \EndIf
     \EndIf
  \Else
     \State \textbf{let} $a' \assigns \text{getReplace}(a)$
     \If {$!isPWait(a') \AndAnd \forall c', !isCWait(c', a')$}
        \State send$(\One{EvictReq}{a'})$
        \State \textcolor{trig-color}{$whoTrigger(a) \assigns Evict$}
        \State \textcolor{trig-color}{$childTrigger(a) \assigns c$}
        \State $replacingAddr(a') \assigns a$
     \EndIf
  \EndIf
\EndRule
\end{algorithmic}
\caption{Handle request from child}
\label{alg:handleCReq}
\end{figure}

\begin{figure}
\begin{algorithmic}
\Rule{handleEvict}{\One{EvictReq}{a}}
   \If {$!isPresent(a) \OrOr \forall c, dir(c, a) = \Inv$}
      \State remove$(\One{EvictReq}{a})$
      \If {$state(a) \neq \Inv$}
         \State send$(\two{RespToP}{a}{\Inv})$
         \If {dataXfer$(state(a), \Inv)$}
            \State send$(\two{DataToP}{a}{data(a)})$
         \EndIf
      \EndIf
      \State $isPresent(a) \assigns \False$
      \State allocLine$(replacingAddr(a))$
      \State \textcolor{trig-color}{\textbf{trigger} handleCReq($childTrigger(a)$)}
   \Else
      \If {$\forall c, !isCWait(c, a)$}
         \ForAll{$c$}
            \If {$dir(c, a) > \Inv$}
               \State $isCWait(c, a) \assigns \True$
               \State $cWait(c, a) \assigns \Inv$
               \State send$(\three{ReqToC}{c}{a}{\Inv})$
            \EndIf
         \EndFor
      \EndIf
   \EndIf
\EndRule
\end{algorithmic}
\caption{Evict a line in address $a$}
\label{alg:handleEvict}
\end{figure}

@@

\begin{figure}
\begin{algorithmic}
\Procedure{allocLine}{$a$}
  \State $isPresent(a) \assigns \True$
  \State \textcolor{trig-color}{$whoTrigger(a) \assigns None$}
  \State $state(a) \assigns \Inv$
  \ForAll {$c$}
     \State $dir(c, a) \assigns \Inv$
     \State $isCWait(c, a) \assigns \False$
  \EndFor
  \State $isPWait(a) \assigns \False$
\EndProcedure
\end{algorithmic}
\caption{Allocate a new cache line for address $a$}
\label{alg:allocLine}
\end{figure}

Each handler will be triggered repeatedly till the corresponding message has
been removed. In order to avoid starvation, the following two conditions must be
maintained:

\begin{enumerate}

\item Once a response is available from any source, the handler for that
response must be triggered eventually.

\item After handling a response (or an evict request) the corresponding handler
must trigger the original handler that requested the received response (or made
an eviction request). The $whoTrigger(a)$ keeps track of the original handler
for each address $a$. In case of a $Child$ trigger, the actual child
corresponding to the original handler is kept track of in $childTrigger(a)$, for
address $a$, in which case $handleCReq$ is triggered for the outstanding request
from that child. In the figures, the pseudo-code corresponding to this is grayed out

\end{enumerate}

The L1 caches, \ie the leaf nodes, do not have any children. Instead, they
handle requests from cores, which resembles handling requests from children. A
store request for a word is essentially a request by the core for write
permissions for the corresponding cache line, and a load request for a word is a
request for read permissions. These permissions can be mapped to the values of
the coherence state.

Because of the absence of children any request from the parent or an eviction
request can be handled without producing more requests. As a result, we do not
need a separate handler for eviction requests. We also do not need to store
$whoTrigger$, since the only trigger that produces more requests, and hence has
to wait for response is a request from the core. We omit the pseudo-code for
handling a request from the core because it is very similar to an instance of
Figure~\ref{alg:handleCReq} with no children.

%Figure~\ref{alg:handleCoreReq} shows the handler for a request from the core,
%which is very similar to a combination of \texttt{handleCReq} and
%\texttt{handleEvict}. The code shows that the cache line is sent back to the
%core for any request, which is not necessary for a store request in an actual
%implementation.
%
%\begin{figure}
%\begin{algorithmic}
%\Rule{handleCoreReq}{\three{ReqFromCore}{a}{x}}
%  \If {$isPresent(a)$}
%     \If {$!isPWait(a)$}
%        \If {$state(a) \ge x$}
%           \State remove$(\three{ReqFromCore}{a}{x})$
%           \State send$(\three{RespToCore}{a}{x}{data(a)})$
%        \ElsIf {$state(a) < x$}
%              \State $isPWait(a) \assigns \True$
%              \State send$(\two{ReqToP}{a}{x})$
%        \EndIf
%     \EndIf
%  \Else
%     \State \textbf{let} $a' \assigns \text{getReplace}(a)$
%     \If {$!isPWait(a')$}
%        \If {$state(a') \neq Inv$}
%           \State send($\two{RespToP}{a'}{Inv}$)
%           \If {dataXfer($state(a'), Inv$)}
%              \State send($\two{DataToP}{a'}{Inv}$)
%           \EndIf
%        \EndIf
%        \State $isPresesnt(a') \assigns \False$
%        \State allocLine($a$)
%        \State $isPWait(a) \assigns \True$
%        \State send$(\two{ReqToP}{a}{x})$
%     \EndIf
%  \EndIf
%\EndRule
%\end{algorithmic}
%\caption{Handle request from core}
%\label{alg:handleCoreReq}
%\end{figure}

@@

### 2012-11-07 paper/isca40format/Introduction.tex
will be removing evict req

\emph{Example 2:} A cache, already having read permission for a line, gets a
store request, and sends an upgrade request for write permission to the
directory controller. Instead of getting a response, the cache may receive a
request from the directory controller asking this cache to give up the read
permission for the line \ie an invalidate request. This could happen if the
directory controller received a request for write permission from another cache
earlier. It is not easy to determine the correct behavior of the cache
receiving the invalidate request -- whether it should ignore the invalidate
request and wait till it receives a response from the directory controller for
its request, or if it should invalidate its line. If it ignores the invalidate
request, is there any guarantee that it would receive a response for its
request for write permission, or would it deadlock?

\emph{Example 3:} A cache receives an invalidate request from the directory
controller for a line that is not present in the cache. Is such a request even
possible? The directory does not have access to the up-to-date state of the
cache, and so can potentially make this request. How should a cache handle this
request; should it drop such a request, or respond saying that it does not have
the line?

>>>>

\noindent \emph{Example 2:} A cache, already having read permission for a line,
gets a store request, and sends an upgrade request for write permission to the
directory controller. Instead of getting a response, the cache may receive a
request from the directory controller asking this cache to give up the read
permission for the line \ie an invalidate request. This could happen if the
directory controller received a request for write permission from another cache
earlier. It is not easy to determine the correct behavior of the cache
receiving the invalidate request -- whether it should ignore the invalidate
request and wait till it receives a response from the directory controller for
its request, or if it should invalidate its line. If it ignores the invalidate
request, is there any guarantee that it would receive a response for its
request for write permission, or would it deadlock?

\noindent \emph{Example 3:} A cache receives an invalidate request from the
directory controller for a line that is not present in the cache. Is such a
request even possible? The directory does not have access to the up-to-date
state of the cache, and so can potentially make this request. How should a
cache handle this request; should it drop such a request, or respond saying
that it does not have the line?
